---
title: 'Will You Be A Halloween Hero Or The Neighborhood Dud?'
author: Shannon Ellis
date: '2018-10-06'
slug: data-driven-decision-making-halloween-candy-edition
categories:
  - R
tags:
  - analysis
  - candy
output: 
  code_folding: hide
---


Now that we're knee-deep into October, leaves are falling, temperatures are cooling, PSLs are flowing from Starbucks like water, and Halloween candy has completely taken over seasonal aisles in grocery stores across the country, it's __time for you to decide what candy to buy for trick-or-treaters__. 

Sure, you could take an economic approach and just get whatever's cheapest (if you're a grad student, I *fully* support this route), or you could just as well take an approach to ___give the people what they want___! 

To help you accomplish this, you all must know [David Ng](http://www.scq.ubc.ca/author/davidng/) has provided us all with a *wealth* of critical data. [Candy hierarchy data are available](http://www.scq.ubc.ca/so-much-candy-data-seriously/) from the last four years. These survey data rank *lots* of different Halloween goodies by preference. Survey respondents let us know which items bring them "JOY," which cause them "DESPAIR," and which are just "MEH." Results from the 2017 survey are here in this [awesome graphic](https://boingboing.net/2017/10/30/the-2017-halloween-candy-hiera.html).

It's important you become aware of these data *now* before you go out and purchase the goodies you'll hand out this year. In this analysis, we'll look at __candy preference trends over time__  , so that you can be *extra* confident in your Halloween Candy purchase. It's up to you whether you want to be a ___Halloween hero___ or the ___biggest dud on your block___.  

```{r setup, include=FALSE}
library(janitor)
library(readr)
library(stringr)
library(reshape2)
library(readxl)
library(ggplot2)
library(dplyr)
```

## The Data

The data are [all available](http://www.scq.ubc.ca/so-much-candy-data-seriously/); however, for the puproses of this post, it's most important you know that each year survey participants are asked how different Halloween goodies make them feel( "MEH", "JOY", or "DESPAIR"). , the __Net Feelies__ metric is calculated as : __(= #JOY - #DESPAIR)__.

Thanks to Jenny Bryan for making me aware of these data with her tweet!

```{r echo=FALSE}
blogdown::shortcode('tweet', '923302921839656960')
```

### Time to Wrangle

There are a few *quirks* in these data. CSVs aren't available for every year. Column names differ slightly. The 2014 and 2015 surveys didn't have a "MEH" option. And, the raw data aren't available for 2014 (just the summarized data). Thus, I decided to make all the data look like the 2014 data, calculating *net feelies* in each dataset from the raw data. 

_Note_: If you saw me [present this](https://docs.google.com/presentation/d/1pDdmLAvg6wEwuz6XrM1FMddXv60xnojmBtmOZyIFqAk/edit?usp=sharing) at R-Ladies Baltimore Launch and heard me say that I'd clean up this code before posting it online, I straight *lied* to you. It's _still_ not pretty. Life happens.

#### 2017

```{r 2017, warning=FALSE, message=FALSE} 
## create directory for data 
if(!dir.exists('candy_data')){
  dir.create('candy_data')
}

## download data from the Internet
if(!file.exists('candy_data/candyhierarchy2017.csv')) {
  system('wget http://www.scq.ubc.ca/wp-content/uploads/2017/10/candyhierarchy2017.csv -P candy_data')
}

## read data into R
candy17 <- read_csv("candy_data/candyhierarchy2017.csv")
names(candy17) <- names(candy17) %>% str_replace_all("\\\\xd5", "")
candy17 <- as.data.frame(candy17)

## determine which columsn we're interested in
cols_to_summarize <- grep("Q6", colnames(candy17)) 

## let's clean up those columns
df <- candy17[,cols_to_summarize] %>% clean_names() 
colnames(df) <- gsub("q6_","",colnames(df))

## function to summarize despair and joy
summary_function <- function(x){
  a <- tabyl(x) %>% 
    melt(id="x") %>% 
    select(value) %>% 
    t() 
  return(a)
}

## get the names of all the columns we'll have in our output object
## who knows why I did it this way ¯\_(ツ)_/¯
x <- df[,2]
nameit <- tabyl(x) %>%
  melt(id="x") %>%
  mutate(category = paste0(x,"_",variable)) %>%
  select(category)

## run function to summarize joy and despair
output17 <- as.data.frame(t(apply(df, 2, summary_function)))
colnames(output17) <- nameit$category

## calculate net feelies
feelies17 = as.data.frame(as.numeric(output17$JOY_n - output17$DESPAIR_n))
rownames(feelies17) <- rownames(output17)
```

##### Data Summary

In 2017, observations were recorded from `r nrow(candy17)` individuals and for `r nrow(feelies17)` different Halloween items.  

#### 2016

```{r 2016, warning=FALSE, message=FALSE}
## download data from the Internet
if(!file.exists('candy_data/BOING-BOING-CANDY-HIERARCHY-2016-SURVEY-Responses.xlsx')) {
  system('wget https://www.scq.ubc.ca/wp-content/uploads/2016/10/BOING-BOING-CANDY-HIERARCHY-2016-SURVEY-Responses.xlsx -P candy_data')
}

## read data into R
candy16 <- read_excel("candy_data/BOING-BOING-CANDY-HIERARCHY-2016-SURVEY-Responses.xlsx") 
candy16 <- as.data.frame(candy16)
candy16 <- candy16[, !apply(is.na(candy16), 2, all)]

## get the columns you're interested in
cols_to_summarize <- grep("^\\[", colnames(candy16)) 
df16 <- candy16[,cols_to_summarize] %>% clean_names()

## use that function from above to summarize joy and despair for each item
output16 <- as.data.frame(t(apply(df16, 2, summary_function)))
colnames(output16) <- nameit$category

## calculate net_feeles
feelies16 = as.data.frame(as.numeric(output16$JOY_n - output16$DESPAIR_n))
rownames(feelies16) <- rownames(output16)

```

##### Data Summary

In 2016, observations were recorded from `r nrow(candy16)` individuals and for `r nrow(feelies16)` different Halloween items. 

#### 2015

```{r 2015, warning=FALSE, message=FALSE}
## download data from the Internet
if(!file.exists('candy_data/CANDY-HIERARCHY-2015-SURVEY-Responses.xlsx')) {
  system('wget https://www.scq.ubc.ca/wp-content/uploads/2015/10/CANDY-HIERARCHY-2015-SURVEY-Responses.xlsx -P candy_data')
}

## read data into R
candy15 <- read_excel("candy_data/CANDY-HIERARCHY-2015-SURVEY-Responses.xlsx")

## get the columns you're interested in
cols_to_summarize <- grep("^\\[", colnames(candy15)) 
df15 <- as.data.frame(candy15[,cols_to_summarize])%>% clean_names()
colnames(df15)<-gsub("x_","",colnames(df15))

## see note in section above
x <- df15[,2]
nameit <- tabyl(x) %>%
  melt(id="x") %>%
  mutate(category = paste0(x,"_",variable)) %>%
  select(category)

## use that function from above to summarize joy and despair for each item
output15 <- as.data.frame(t(apply(df15, 2, summary_function)))
colnames(output15) <- nameit$category

## calculate net feelies
feelies15 = as.data.frame(as.numeric(output15$JOY_n - output15$DESPAIR_n))
rownames(feelies15) <- rownames(output15)
```

##### Data Summary

In 2015, observations were recorded from `r nrow(candy15)` individuals and for `r nrow(feelies15)` different Halloween items.  

#### 2014

```{r 2014, warning=FALSE, message=FALSE}
## download data from the Internet
if(!file.exists('candy_data/CANDYDATA.xlsx')) {
  system('wget http://www.scq.ubc.ca/wp-content/uploads/2014/10/CANDYDATA.xlsx -P candy_data')
} 

## read data into R & clean it up a little
candy14 <- read_excel("candy_data/CANDYDATA.xlsx") %>% clean_names %>%
  as.data.frame %>%
  filter(!is.na(item))
rownames(candy14) <- candy14$item
candy14 <- as.data.frame(t(candy14))
candy14<-clean_names(candy14)
colnames(candy14)<-gsub("x100","100",colnames(candy14))

candy14 <- as.data.frame(t(candy14))

## net feelies have already been calculated
feelies14 <- as.data.frame(as.numeric(as.character(candy14$net_feelies)))
rownames(feelies14) <- rownames(candy14)
```

##### Data Summary

In 2014, observations were recorded from 1286 individuals and for `r nrow(feelies14)` different Halloween items.  

### Merging & Scaling

To complete this analysis, all four years of "feelies" data are combined into a single data frame. They are then scaled within each year so that the most favorable candy (brings the most relative "JOY") receives a score of +1 while the least favorable (brings the most relative "DESPAIR") receives a score of -1. 

```{r data-merge-all}
## merge all these data together
combined_all <- merge(feelies14,feelies15,by = "row.names", all = TRUE) %>%
  merge(.,feelies16,by.x = "Row.names",by.y = "row.names", all = TRUE) %>%
  merge(.,feelies17,by.x = "Row.names",by.y = "row.names", all=TRUE)
combined_all <- as.data.frame(combined_all)
colnames(combined_all) <- c("row.names","x2014","x2015","x2016","x2017")

## scale all values between -1 and 1 within a year
scalefunc <- function(x){
  minx = min(x, na.rm = TRUE)
  maxx = max(x, na.rm = TRUE)
  val= 2*((x - minx)/(maxx - minx))-1
  return(val)
} 
out_all <- as.data.frame(apply(combined_all[,2:5],2,scalefunc))
out_all$row.names <- as.character(combined_all$row.names)

## calculate a few things we may use to order the graph
## calculate sum, normalized by number of years question was asked
out_all$sum <- rowSums(out_all[,1:4],na.rm=TRUE)
isNA <- function(x){sum(is.na(x))}
out_all$isNA <- apply(out_all,1,isNA)
out_all$sum_NA <- out_all$sum/(4-out_all$isNA) 
out_all <- out_all %>% arrange(sum_NA)

```

## Give the People What They Want!

What you've been waiting for - the results! Each year is a different color point on the graph (see legend on the plot). Rows are ordered such that items with the most positive net feelies are listed at the top of the graphic (*Halloween Hero* territory!) and those that consistently cause the most despair (*Neighborhood Dud* Land) are at the bottom. 

```{r data-viz-all, warning=FALSE, fig.height=35, fig.width = 7}
## tidy data format!
data_all <- melt(out_all, id.vars='row.names', 
                 variable.name='Year', value.name = "net_feelies") %>%
  mutate(item_name = gsub("_", " ",row.names))

## let's plot!
data_all %>%
  filter(Year=="x2014" | Year=="x2015" | Year=="x2016" | Year=="x2017") %>%
  ggplot(., aes(x = factor(item_name, level = unique(item_name)), 
                y = net_feelies, 
                color = Year)) +
  geom_point(size = 4) +
  scale_colour_manual(
    labels = c("x2014" = "2014", 
               "x2015" = "2015", 
               "x2016" = "2016", 
               "x2017" = "2017"),
    values = c("x2014" = "#FFEE4A", 
               "x2015" = "#FE9600", 
               "x2016" = "gray48",
               "x2017" = "#77477E")) +
  geom_segment(aes(x = item_name, 
                   xend = item_name, 
                   y = -1, 
                   yend = 1), 
               linetype = "dashed", 
               size = 0.1) +   # Draw dashed lines
  scale_x_discrete(labels = function(y) str_wrap(y, width = 47)) +
  labs(title ="Candy Preferences Over Time", 
       subtitle ="So Much Candy Survey", 
       caption ="source: http://www.scq.ubc.ca/so-much-candy-data-seriously/",
       x="",
       y="net feelies (scaled)") +  
  coord_flip() +
  theme(legend.position = c(-0.7, 0.9),
        legend.text = element_text(size = 16),
        axis.text.x = element_text(size = 16, hjust = 0.5),
        axis.text.y = element_text(size = 11),
        axis.title.y = element_text(size = 14),
        legend.title = element_text(size = 20),
        plot.title = element_text(size = 24, hjust = 1.2, color = "#77477E"),
        plot.subtitle = element_text(size = 16, hjust = -0.45, color = "#FE9600"))
```

## Take-Away Messages

When I look at this plot, I conclude that candy preferences do *not* vary wildly from year to year. That's good for you! What the people have liked in the past is likely to be what they'll like this year!

For items in the middle of the plot where scaled net feelies appear to differ more from year to year, I want to remind you that "MEH" was not an option in 2014 and 2015, suggesting that when not given the "MEH" choice, people tend to say "DESPAIR."

### Halloween Hero Territory

Probably unsurprisingly, __full-sized candy bars are the bee's knees__ on Halloween. But, we're not all made of money. Fortunately, __Reese's peanut butter cups,  Kit Kat bars, Twix, and Snickers__ consistently bring almost as much joy as full-sized candy bars. So, if you want to maximize the joy your treats bring your trick-or-treaters while not breaking the bank, opt for one of these classics!


### Neighborhood Dud Land

On the other end, if you want to be known as the neighborhood's worst stop for Halloween treats, give out __broken glow sticks__! Similarly, __raisins, dental paraphernalia, and sidewalk chalk__ are also likely to land you on the worst-candy neighborhood list.

### The Sleepers

If you're driven to be _unique_,  __Dove bars, Lindt truffles, and Mars__ are less popular (as determined by my many years of trick-or-treating experience) but bring joy the masses (as proven by these data)!

### Personal Hot-Take

Personally, I think __Kit Kats are _way_ too high__ on this list and are a completely mediocre (*at best*) candy.

Similarly, I take offense to the fact that the respondents' preferences are _so_ heavily skewed to chocolate candy. Don't get me wrong, chocolate is great. But that said, the first non-chocolate item to show up is _chardonnay_. Chardonnay!?!?! While _delicious_, is certainly *not* a Halloween candy to be given out to minors. I'm here to lobby for the joy brought to me by gummy candy! __Swedish fish! Sour Patch Kids! Good N Plenty!__ These candies are getting the damn _shaft_ on this list.

Finally, if you ever find yourself with a kale smooothie on your hands that you're not interested in drinking, I'd be happy to take it off your hands.


