---
title: 'Teaching Opportunities: Speech Speed'
author: "Shannon Ellis"
date: '2021-06-10'
slug: speech-speed
categories:
  - education
tags:
  - analysis
  - classroom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)

library(googlesheets4)
library(tidyverse)
```



```{r}
read_data <- function(sheet, quarter){
  read_sheet(sheet) %>% 
  mutate(quarter = quarter) %>% 
  distinct(., `Email Address`, .keep_all = TRUE) %>% # remove duplicate responses
  rename(speech = `Please respond to the following statements considering your *experience in COGS 108* specifically: [Prof Ellis speaks too quickly.]`) %>% 
  select(quarter, speech)
}

fa20 <- read_data(sheet = '1qs_msoEfd1gJb8tHESNGZSIZfMyS1Uv57e7PMxbFkv4', 
                  quarter = 'fa20') 
wi21 <- read_data(sheet = '1uFFGE_G252AnGA0ZmKXgwV08mrIGm1zlHp6mQck5phs', 
                  quarter = 'wi21') 
sp21 <- read_data(sheet = '1GK7bD6gGbKKB6KMQjGKIlfAK8bKAgI23DfgbT4sQGWw', 
                  quarter = 'sp21') 
                  
toorder = c("Strongly Agree", "Somewhat Agree", "Neutral", "Somewhat Disagree", "Strongly Disagree")

df <- bind_rows(fa20, wi21, sp21) %>% 
  filter(!str_detect(speech, ',')) %>% 
  mutate(speech = fct_relevel(speech, toorder),
         quarter = fct_relevel(quarter, c("sp21", "wi21", "fa20")))

summ <- summarized_table <- df %>% 
  group_by(quarter, speech) %>%
  summarise(n = n()) %>%
  mutate(freq = n / sum(n),
         perc = round(freq*100, 2))

```

When I would give talks as a postdoc and as I was applying to academic jobs, I would often joke with audiences that I come from a part of the country where people speak really fast (Philadelphia), and even among Philadelphians, I'm regarded as a fast talker. In fact, when interviewing for my current role at UC San Diego, our former chair commented on how quickly I speak a handful of times during my interview, negotiations, and the eventual offer.

During my interview and many times since, I've asked many people about the speed at which I speak, and while most have agreed I speak more quickly than average, many have also suggested that I'm clearer than most in my speech, and thus they don't mind the speed. Before arriving at UC San Diego, I wasn't too concerned and would try to slow down without focusing on it too much. However, I'm now a teaching professor at a school with a large international population (often ~40% of my students are international students). And, I know that when I try to listen to fast-talking Spanish speakers (Spanish is my non-native language), I struggle. Now, my students for whom English is not their native language, are largely better in their non-native language than I am in mine, but nevertheless I have been concerned as to whether I'm putting a subset of my students at a disadvantage because of the speed at which I talk. 

In conversations with many international students over the past few years, I've asked them about the speed at which I talk, and they have all said it's fine....but let's be honest, I'm their professor (um, hello power dynamic) and my students are just generally really kind people. So, I wanted to get a sense of how large an issue this was for my students.

### Approach #1: Anonymous Feedback

While students have avenues to ask questions in my courses in a manner in which they are anonymous to their peers, they are not anonymous to myself or the instructional staff. Therefore, I have a Google Form that students can use to provide *truly* anonymous feedback. To introduce this form on the first day of class each term and to prime them to the topic of my speech speed, I say something along the lines of: 

> "This form is the only place in which you'll be *truly* anonymous in this course, and I am the only person who has access to what you share here. If you ever have thoughts or feedback about how I can improve your educational experience and you'd rather not have your name attached to the comment, please feel free to use this. Note that if this form is used in a manner other than its specific intention, I can take it down. For example, if you don't like the sound of my voice, that is *not* appropriate feedback for the anonymous form, as I cannot change the sound of my voice. *However*, if the *speed* at which I'm speaking is too fast for you, that *would* be appropriate feedback, as that is something that I can change and is something that could affect your educational experience."

Since opening this form in January 2019, I have taught more than 5,000 students and have received the following four comments regarding my speech speed: 

> *It is very hard to keep up with your online lectures as you go so fast that I have to constantly pause to keep up. Although it's nice that I'm able to pause, it makes the time I allot for the class twice of what you suggested.*

> *Would you please speak more slowly in the lecture class? Since you spoke so fast, it was kind of hard for me to follow.*

> *I think it would be helpful if you could talk a little slower in the videos please, because i have to rewind to be able to follow the material. Thanks!*

> *Professor Ellis, if you could please speak a little slower, I'd truly appreciate it. English is not my first language, so I sometimes have a hard time understanding/following you in lecture.*

Each time I received one of these comments, I simultaneously appreciate the student taking the time to let me know and feel I'm falling short of my goals as an educator. While even one student struggling to understand me in lecture is too many, I was curious how widespread this issue was. 

### Approach #2: Direct Survey

In COGS 108 Data Science in Practice (the course where I get really excited and speak too quickly and from which these comments have all come) in Fall 2020, I started including a Likert-scale question on the post-course survey that I ask students to fill out each quarter. The results summarized across three quarters are below: 


```{r}
ggplot(df, aes(x=fct_relevel(speech, toorder), fill=speech)) +
  geom_bar() +
  labs(title = "Prof Ellis speaks too quickly.",
       y = "No. of Responses") +
  scale_fill_manual(values = c( "#1F497C", "#1F497C", "#BFBFBF", "#404040", "#404040")) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  theme_classic(base_size = 14) +
  theme(legend.position="none", 
        plot.title.position = "plot",
        axis.title.x = element_blank())
```

Here, I'm most concerned with the blue bars - students who say they "Strongly Agree" or "Somewhat Agree" to the statement "Prof Ellis speaks too quickly." It was reassuring that it does not seems to be an overwhelming issue for my students at large, but I still wanted to address it. I figured a small improvement each quarter would be my goal, and I have now asked this question in COGS 108 for three consecutive quarters: Fall 2020, Winter 2021, and Spring 2021, enabling me to determine if I've met my goal. 

```{r}
p <- ggplot(summ, aes(x = quarter, y=perc, fill=speech)) +
  geom_bar(position="fill", stat="identity", colour = "white") +
  labs(title = "Prof Ellis speaks too quickly.",
       y = "% of Responses") +
  scale_fill_manual(values = c( "#1F497C", "#1F497C", "#BFBFBF", "#404040", "#404040")) + 
  scale_colour_manual(values = "white") + 
  scale_y_continuous(labels = scales::percent) + 
  coord_flip() + 
  theme_classic(base_size = 14) + 
  theme(plot.title.position = "plot",
        axis.title.y = element_blank(),
        legend.title = element_blank())
p

vals <- summ %>% filter(str_detect(speech, "Agree"))
```


While I did improve slightly from the Fall 2020 to Winter 2021 quarter (`r round(vals %>% filter(quarter=="fa20") %>% ungroup() %>% select(perc) %>% sum(), 1)`% to `r round(vals %>% filter(quarter=="wi21") %>% ungroup() %>% select(perc) %>% sum(), 1)`% of students saying they somewhat or strongly agree that I speak too quickly), this increased slightly in the Spring 2021 quarter (`r round(vals %>% filter(quarter=="sp21") %>% ungroup() %>% select(perc) %>% sum(), 1)`%).


### Interpretation

All three of these quarters were taught remotely due to the COVID-19 pandemic. However, the way I delivered the course differed from one quarter to the next. Specifically, during Fall 2020, lecture was delivered asynchronously via short pre-recorded videos. Given that these are, by definition, not interactive, I would not be surprised if, while recording, I sped through videos unconsciously, even when I was really trying to speak at a reasonable pace. Then, in the Winter 2021 quarter (lowest % of students indicating an issue with speech speed), I both delivered lectures synchronously *and* I had a particularly active and attentive class of students who asked more questions than typical during lecture. Given that I'd clarify more during class while answering questions, this led me to get "behind" in the Winter quarter and cover less material overall.  I have a feeling this likely led some students to feel less rushed to understand during lecture than in other quarters. Spring 2021 was delivered in a similar structure as in the Winter 2021; however, I removed and consolidated some content to get through all necessary material without rushing/speaking more quickly. However, given results here, I'm not sure I accomplished the latter. Considering this, the next time I teach COGS 108 (likely the 2022-2023 academic year), I'm going to revisit the amount of content I teach, figure out what can and should be cut, reworked, or reorganized to minimize the need to rush and continue to find ways to control my speech speed when we're back in the classroom. Finally, while I'm glad my speech speed is not negatively affecting a majority of my students, I'm going to continue work to ensure that students find my courses clear and my speech understandable.

