---
title: GenAI In Intro Programming (a follow-up)
author: Package Build
date: '2026-07-23'
slug: genai-in-intro-programming
categories:
  - education
tags:
  - ai
  - classroom
  - postmortem
---


```{r, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

library(tidyverse)
library(googlesheets4)
library(kableExtra)
library(patchwork)

options(gargle_oauth_email = "sellis@ucsd.edu")
options(knitr.table.format = "html") 
```

## tl;dr Summary
- 93% of students report using GenAI for schoolwork (+21% from 2023)
- Frequency of use has increased, with 26% of students reporting using GenAI "most days" (+17% from 2023)
- Students conceive of GenAI as a personalized tutor (rather than just a "tool")

## Introduction 


In 2023 I summarized [GenAI usage in COGS 18 Introduction to Python](https://shannon-ellis.com/blog/your-students-are-using-chatgpt/), with 72% of students reporting having used GenAI for school tasks with few using it frequently. Three years later, and with a course redesign partially targeted at helping students to use GenAI to effectively support their learning, I was curious how usage trends and sentiments had changed.

```{r cache=TRUE}
sp23 <- read_sheet('1BN11yA4Xc1P5Hgaj0BIXKiwvsf1nLXEcq6a963ZBhyQ')
sp26 <- read_sheet('1uskvhAXJvmBmZZydQYMda802kfE3suJmh1kYxo1GfVo')
```

## Frequency of use

In Spring 2026, 439 students completed the post-course survey with **93% reporting using GenAI for schoolwork**. This is a **21% increase from Spring 2023**. Frequency of use has also shifted, with more students reporting frequent use than they did a few years ago. (**26% reporting using GenAI most days compared to 9% in 2023**):

```{r}
library(tidyverse)

# 1. Select and standardize the question column
sp23_clean <- sp23 |>
  transmute(
    quarter = "Sp23",
    response = `How often do you use ChatGPT (or a similar technology) for schoolwork?`
  )

sp26_clean <- sp26 |>
  transmute(
    quarter = "Sp26",
    response = `How often do you use GenAI for schoolwork (including COGS 18 and other courses)?`
  )

# 2. Combine the two quarters
chatgpt_use <- bind_rows(sp23_clean, sp26_clean)

# 3. Calculate response proportions within each quarter
plot_data <- chatgpt_use |>
  filter(!is.na(response)) |>
  count(quarter, response) |>
  group_by(quarter) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

freq_plot <- ggplot(plot_data, aes(x = fct_relevel(response, "Frequently (most days)", after=3), y = proportion, fill = quarter)) +
  geom_col(position = "dodge") +
   geom_text(
    aes(label = scales::percent(proportion, accuracy = 1)),
    position = position_dodge(width = 0.9),
    vjust = -0.3
  ) +
  scale_fill_manual(
    values = c(
      "Sp23" = "#f1a340",
      "Sp26" = "#998ec3")) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  labs(
    x = "Frequency of GenAI use for schoolwork",
    y = "Proportion of respondents",
    fill = "Quarter"
  ) +
  theme_classic(base_size = 14) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 12)) +
  theme(plot.title.position = "plot",
        axis.title = element_blank(),
        axis.ticks = element_blank()) 
```

```{r, fig.height=2, fig.show=TRUE}
print(freq_plot)
```


## Technology Holdouts

Earler in the term I had asked students who've chosen not to use GenAI technology why they've made that choice. Student responses revolved around:

- **wanting to ensure they're learning** ("*I do not believe in it. If I do bad at least it is on my own accord.*" and "*I want to actually learn the material and not rely on AI to learn the material, and just mirror what it does.*")
- **environmental concern**s ("*I'm personally against using AI due to its environmental impacts*" and "*I don't agree with how AI information is obtained. I'm also highly against its environmental impact.*")
- **ethical concerns** ("*I believe that AI is causing or contributing to very serious issues worldwide and so I don't use it in any of my classes or in my free time*" and "*I haven't used GenAI because I don't agree with the ethical issues surrounding. I also find that I learn more when I ask a person to explain it to me.*")
- **not needing it** ("*Mainly personal preference and also generally don't need it.*")

## How have students perceptions of benefits and limitations changed? 

In both iterations students were asked how GenAI has helped or limited students in their abilities to learn. Summarizing these responses, students report that **GenAI was helpful**: 

- acting as a **personal tutor**: students highlight that it adapts to their learning style, has unlimited patience, does not judge, available at all hours
- in debugging (mentioned more often than conceptual help)
- **generating practice problems**: students were explicitly encouraged to do this and shown how effective this can be during Spring 2026
- because they **prompted not to give answers** ("DO NOT ANSWER JUST EXPLAIN").

When asked how  "How do you think GenAI has limited your ability to learn in COGS 18?" student responses centered around: 

- **over-reliance** (students' top concern): students commented on "learned helplessness," "offloading mental capacity," "limited cognitive friction," and "hijack my thinking."
- **substituting for humans**: students lamented its use displacing social interaction in office hours, labs, and peer interactions
- **fast-fading understandin**g: with students spending less time sitting with content, felt knowledge didn't stick.

While full responses from Spring 2023 are [included in the previous post](https://shannon-ellis.com/blog/your-students-are-using-chatgpt/) full student responses from Spring 2026 are included below:

<details><summary>Helped me learn (Spring 2026) </summary>

```{r, results="asis"}
sp26 %>%
  select(`(optional) How do you think GenAI has helped you learn in COGS 18?`) %>% 
  filter(!is.na(`(optional) How do you think GenAI has helped you learn in COGS 18?`)) %>% 
  kbl() %>%
  kable_styling(full_width = F) %>% 
  scroll_box(width = "700px", height = "400px")
```
</details>


<details><summary>Limited learning (Spring 2026) </summary>

```{r, results="asis"}
sp26 %>%
  select(`(optional) How do you think GenAI has limited your ability to learn in COGS 18?`) %>% 
  filter(!is.na(`(optional) How do you think GenAI has limited your ability to learn in COGS 18?`)) %>% 
  kbl() %>%
  kable_styling(full_width = F) %>% 
  scroll_box(width = "700px", height = "400px")
```
</details>

Notable shifts from Spring 2023 include: 

- students are no longer interacting with the technology for the first time; safer to assume familiarity
- skepticism in accuracy has decreased as the technology has developed
- students previously framed GenAI as a tool and now are more likely to frame it as a tutor or assistant
- students now (but did not previously state they) have concerns about the technology displacing humans and/or the need for human interaction 


## Discussion

Given that I explicitly teach students how to learn alongside GenAI, it's likely not surprising that **I'm pleased to see more students familiar with how to use GenAI in their studies**. However, of course, I do value students environmental and ethical objections and have concerns on student over-reliance. However, I personally think that, for a course where GenAI can assist students in learning, the solution to avoid over-reliance and ensure learning is for instructors of a given course to design their course accordingly (i.e. through explicit instruction, oral and/or secure assessments, etc.) rather than to ban its use.